Basics of Hardware Cache Optimization

### 2017-06-13 paper.tex
Put datavisualization configuration into new file
Expand content, revise, tweak
Draft section on locality, tweak stuff
Tweak, fix typo
Add program comparing std::list and std::vector
Write a bit about the EMM; revise and tweak stuff

@@

% Stack for generating tables from CSV files.  Loading is done by [pgfplotstable].  It can
% also round, format and post-process data.  See [1].
% [siunitx] adds the `S` column type for advanced aligning in tables like around the
% decimal marker.  Getting this to work is a bit tricky; see [2].  There is some support
% for aligning around the decimal marker in [pgfplotstable] (see [3]), but I couldn't get
% it to work.  One can also right-align numbers relatively to each other but still center
% the column as a whole with [siunitx] (see [4]).
% [pgfplotstable]: https://ctan.org/pkg/pgfplotstable
% [siunitx]: https://ctan.org/pkg/siunitx
% [1]: https://tex.stackexchange.com/q/146716
% [2]: https://tex.stackexchange.com/q/131081
% [3]: https://tex.stackexchange.com/q/276485
% [4]: https://tex.stackexchange.com/q/9203
\usepackage{pgfplotstable, siunitx}
% These packages are recommended by [pgfplotstable].  The booktabs package adds the
% `\toprule`, `\midrule`, and `\bottomrule` commands for better looking tables.  See [5].
\usepackage{booktabs, array, colortbl}
% [booktabs]: https://ctan.org/pkg/booktabs
% [5]: https://en.wikibooks.org/wiki/LaTeX/Tables#Professional_tables

@@

% See <https://tex.stackexchange.com/q/10684/78512>.  TODO.
\usepackage{enumitem}
\setlist{noitemsep}

@@

\usepackage{tikz}
\usetikzlibrary{datavisualization}
% \usetikzlibrary{datavisualization.formats.functions}

@@

\begin{abstract}
   % What?
   Typical present-day CPUs have two or more levels of caches.  This \article{} presents
   the basic techniques used to optimize program performance based on knowledge about how
   these hardware caches function.

@@

\input{tex/intro}
\input{tex/motivation}
% \section{Hardware Caches} % Cache, Structure of Caches
\input{tex/operation}
\input{tex/types}
\input{tex/key-concepts}
\section{Example: \texttt{std::vector} vs. \texttt{std::list}}
\label{sec:vvl}

% \section{Cache Replacement Policies} % Eviction Policies/Strategies/Algorithms
% https://en.wikipedia.org/wiki/Cache_replacement_policies

\section{External Memory Model}
The \gls{emm} is a widely used
% TODO: is it?
extension of the \gls{ram} model.

\section{Cache-Oblivious Algorithms} % Cache-Oblivious Model

\section{Parallel Computing}

\subsection{Cache Coherence} % Coherence or coherency?
% https://en.wikipedia.org/wiki/Cache_coherence
% Dirty.  Invalid.

\subsection{False Sharing}

>>>>

% `$\nicefrac{a}{b}$` looks better than `$a/b$` in my opinion.  I also think it looks
% better than `$\sfrac{a}{b}` from the more recent [xfrac][5] package.  See [3], [4], and
% [5].
%
% [1]: https://www.ctan.org/pkg/nicefrac
% [2]: https://www.ctan.org/pkg/xfrac
% [3]: https://tex.stackexchange.com/q/3372
%      "How do I typeset arbitrary fractions like the standard symbol for .5 = ½?"
% [4]: https://tex.stackexchange.com/q/128496 "Elegant fractions in one line"
% [5]: https://tex.stackexchange.com/q/43119 "Improved kerning in fractions?"
\usepackage{nicefrac}

@@

% Stack for generating tables from CSV files.  Loading is done by [pgfplotstable].  It can
% also round, format and post-process data.  See [1].
% [siunitx] adds the `S` column type for advanced aligning in tables like around the
% decimal marker.  Getting this to work is a bit tricky; see [2].  There is some support
% for aligning around the decimal marker in [pgfplotstable] (see [3]), but I couldn't get
% it to work.  One can also right-align numbers relatively to each other but still center
% the column as a whole with [siunitx] (see [4]).
% [pgfplotstable]: https://ctan.org/pkg/pgfplotstable
% [siunitx]: https://ctan.org/pkg/siunitx
% [1]: https://tex.stackexchange.com/q/146716
% [2]: https://tex.stackexchange.com/q/131081
% [3]: https://tex.stackexchange.com/q/276485
% [4]: https://tex.stackexchange.com/q/9203
\usepackage{pgfplotstable}
\usepackage[binary-units]{siunitx} % Load `\kibi` and `\mebi` etc. (`binary-units`).
% These packages are recommended by [pgfplotstable].  The booktabs package adds the
% `\toprule`, `\midrule`, and `\bottomrule` commands for better looking tables.  See [5].
\usepackage{booktabs, array, colortbl}
% [booktabs]: https://ctan.org/pkg/booktabs
% [5]: https://en.wikibooks.org/wiki/LaTeX/Tables#Professional_tables

% This gets loaded by the `pgfplotstable` sub-package as far as I know.
%
% \usepackage{pgfplots}
%
% It then suggests:
%
%    Package pgfplots Warning: running in backwards compatibility mode (unsuitable t
%    ick labels; missing features). Consider writing \pgfplotsset{compat=1.14} into
%    your preamble
%
% So I guess I'll go that.
\pgfplotsset{compat=1.14}

@@

% See <https://tex.stackexchange.com/q/10684/78512>.  TODO.
\usepackage[inline]{enumitem}
\setlist{noitemsep}

% \usepackage{tasks}

@@

\usepackage{tikz}
\input{tex/datavis.tex}

@@

\begin{abstract}
   % What?
   Typical present-day CPUs have two or more levels of caches.
   \alts[3]{
      {This \article{} presents the basic techniques used to optimize program performance
      based on knowledge about how these hardware caches function.},
      {This \article{} explains their key architectural properties with the intend of
      \alts{providing, giving} basic insight on possible program optimizations.},
      {This \article{} provides basic insight into their operation and presents key
      architectural properties which suggest possible program optimizations.},
   }

@@

\input{tex/intro}
\input{tex/motivation}
% \section{Hardware Caches} % Cache, Structure of Caches
\input{tex/operation}
\input{tex/types}
\input{tex/key-concepts}
\input{tex/list-vs-vector.tex}
\label{sec:lvv}
% \section{Cache Replacement Policies} % Eviction Policies/Strategies/Algorithms
% https://en.wikipedia.org/wiki/Cache_replacement_policies
\input{tex/emm.tex}
\section{Parallel Computing}
\subsection{Cache Coherence} % Coherence or coherency?
% https://en.wikipedia.org/wiki/Cache_coherence
% Dirty.  Invalid.
\subsection{False Sharing}

### 2017-06-13 tex/list-vs-vector.tex
Write a bit about the EMM; revise and tweak stuff

% \noindent
% \alts{{The result is that, when}, When}  computing the
% sum \alts{runs, is} \num[round-mode=places, round-precision=0]{\speedup} times faster.
\alts{My, The} result is that computing the sum \alts{runs, is} \num[round-mode=places,
round-precision=0]{\speedup} times faster when using \mintinline{cpp}{std::vector}.  Some
of this difference can certainly be attributed to space overhead of the linked list and
the added indirection, but the more \emph{cache-friendly} memory access pattern of
\mintinline{cpp}{std::vector} is \alts{decisive, key, crucial, vital}~\cite[5\psq]{bigos}.
Using \mintinline{cpp}{std::list} incurs \say{random access to memory} in this
example~\cite[6]{bigos}.

>>>>

% \noindent
% \alts{{The result is that, when}, When}  computing the
% sum \alts{runs, is} \num[round-mode=places, round-precision=0]{\speedup} times faster.
\alts{My, The} result is that computing the sum \alts{runs, is} \num[round-mode=places,
round-precision=0]{\speedup} times faster when using \mintinline{cpp}{std::vector}.  Some
of this difference can \x{certainly} be attributed to space overhead of the linked list
and the added indirection, but the more cache-friendly memory access pattern of
\mintinline{cpp}{std::vector} is \alts{decisive, key, crucial, vital}~\cite[5\psq]{bigos}.
Using \mintinline{cpp}{std::list} incurs \say{random access to memory} in this
example~\cite[6]{bigos}.
